---
title: "Lasso_modeling"
output: html_document
date: "2026-04-24"
---
### Source: https://www.statology.org/lasso-regression-in-r/

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r}
library(readr)
library(dplyr)
library(glmnet)
```

# Load Data
```{r}
df_23_24 <- read_csv("data/raw/2023-2024_NSCH_Topical_CAHMI_DRC.csv")

allfactors_23_24 <- df_23_24 |> 
  select(grades6to11_2324, ChHlthSt_2324, LowBWght_2324, PhysAct_2324, Cond2more_2324, CSHCN_2324, ADHDind_2324, ADHDSevInd_2324, ASDSevInd_2324, bullied_2324, bully_2324, MakeFriend_2324, MEDB10ScrQ5_2324, MedCare_2324, MentHCare_2324, SpEducPln_2324, ReptGrade_2324, povlev4_2324, age5_2324, sex_2324, FoodSit_2324, FoodCash_2324, HousingInstab_2324, famstruct5_2324)

df_23_24_clean <- allfactors_23_24 %>%
  filter(MedCare_2324 %in% c(1,2)) %>%
  filter(MentHCare_2324 %in% c(1,2,3)) %>%
  filter(SpEducPln_2324 %in% c(1,2)) %>%
  filter(ReptGrade_2324 %in% c(1,2)) %>%
  filter(HousingInstab_2324 %in% c(1,2)) %>%
  filter(povlev4_2324 %in% c(1, 2, 3, 4)) %>%
  filter(age5_2324 %in% c(1, 2, 3, 4, 5)) %>%
  filter(sex_2324 %in% c(1, 2)) %>%
  filter(famstruct5_2324 %in% c(1, 2, 3, 4)) %>%
  filter(FoodSit_2324 %in% c(1, 2, 3, 4)) %>%
  filter(FoodCash_2324 %in% c(1, 2, 3))
```

# Establish variables
```{r}
#response variable = grades
y <- allfactors_23_24$grades6to11_2324

#matrix of predictor variables
x <- data.matrix(allfactors_23_24[, c('ChHlthSt_2324', 'LowBWght_2324', 'PhysAct_2324', 'Cond2more_2324', 'CSHCN_2324', 'ADHDind_2324', 'ADHDSevInd_2324', 'ASDSevInd_2324', 'bullied_2324', 'bully_2324', 'MakeFriend_2324', 'MEDB10ScrQ5_2324', 'MedCare_2324', 'MentHCare_2324', 'SpEducPln_2324', 'ReptGrade_2324', 'povlev4_2324', 'age5_2324', 'sex_2324', 'FoodSit_2324', 'FoodCash_2324')])

```

#Create the model

```{r}
# find the least squares regression with variable "lamda"

lasso_model <- cv.glmnet(x, y, alpha = 1)

#find optimal lambda value that minimizes test MSE
best_lambda <- lasso_model$lambda.min
best_lambda
```

#Model Coefficients
### Lasso Regression has the potential to remove predictors by shrinking the coefficient down to 0.
```{r}
bestlasso_model <- glmnet(x, y, alpha = 1, lambda = best_lambda)
coef(bestlasso_model)
```
### Coefficients with "." have their coefficients dropped to 0 as they are not influential enough. This is contradictory to what we have discovered with other data exploration such as ggpairs.


# Making a prediction: what are the likely grades for a child with the characteristics inputted to the variables?
```{r}
#define predictor variables first!

ChHlthSt <-  
LowBWght <- 
PhysAct <- 
Cond2more <-
CSHCN <-
ADHDind <- 
ADHDSevInd<- 
ASDSevInd <- 
bullied <-
bully <-
MakeFriend <-
MEDB10ScrQ5 <- 
MedCare <- 
MentHCare <- 
SpEducPln <- 
ReptGrade <-
povlev4 <-
age5 <-
sex <-
FoodSit <-
FoodCash <-

#define new observation/child
predict_child = matrix(c(ChHlthSt, LowBWght, PhysAct, Cond2more, CSHCN, ADHDind, ADHDSevInd, ASDSevInd, bullied, bully, MakeFriend, MEDB10ScrQ5, MedCare, MentHCare, SpEducPln, ReptGrade, povlev4, age5, sex, FoodSit, FoodCash), nrow=1, ncol=21) 

#use lasso regression model to predict response value
predict(best_model, s = best_lambda, newx = predict_child)
```

# Calculate R Squared
```{r}
y_predicted <- predict(bestlasso_model, s = best_lambda, newx = x)

#find SST and SSE
sst <- sum((y - mean(y))^2)
sse <- sum((y_predicted - y)^2)

#find R-Squared
rsq <- 1 - sse/sst
rsq
```

