---
title: "Logistic Regression Model: Grade Outcomes"
author: "Gabrielle Arrache and Naomi Francis"
output: html_document
date: "`r Sys.Date()`"
---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidymodels)  # modelling framework (recipes, workflows, tuning, metrics)
library(glmnet)      # engine that fits the logistic regression with regularisation
library(vip)         # variable importance plots
library(broom)       # tidy() to extract model coefficients as a data frame
```

---

# Step 1: Prepare the Data

```{r}
#run clean_data.Rmd first
# Convert the outcome to a factor

df_model <- df_23_24_clean |>

# The order of the levels matters here, since “Excellent” will be used as the reference category.
  mutate(grades_2324 = factor(
    grades_2324, 
    levels = c(1, 2, 3, 4),
    labels = c("Excellent", "Good", "Fair", "Poor")
  )) |>
  na.omit()
# Print the class distribution so we can see the imbalance
cat("Grade distribution:\n")
round(prop.table(table(df_model$grades_2324)) * 100, 1)
```

---

# Step 2: Train / Test Split

```{r split}
set.seed(21)  

# Split 70% train / 30% test 
# strata = grades_2324 keeps the same class proportions in both halves, which is important because "Fair" and "Poor" are rare.
split <- initial_split(df_model, prop = 0.70, strata = grades_2324)
train <- training(split)   
test  <- testing(split) 

nrow(train)
nrow(test)
```

---

# Step 3: Compute Class Weights

To fix class imbalance, we are using **class weights**. Rare classes get a
higher weight so the model is penalised more for missing them.

The formula is: `weight = total_rows / count_of_that_class`
This makes the effective contribution of each class equal during training.

```{r class-weights}
# Total number of training rows
n_total <- nrow(train)

# Count how many rows belong to each grade class
class_counts <- table(train$grades_2324)

# Compute the weight for each class
# weight = total / count  -->  rare classes get higher weights
weights_lookup <- n_total / class_counts
print(round(weights_lookup, 2))

# Add the weight as a new column in the training data.
# importance_weights() is the tidymodels way to tell the model
# "pay more attention to rows with higher weights".
train_w <- train |>
  mutate(
    case_wt = importance_weights(
      as.numeric(weights_lookup[as.character(grades_2324)])
    )
  )
```

---

# Step 4: Set Up the Recipe and Model

```{r recipe-and-model}
# recipe() defines how to prepare the data before fitting.
# grades_2324 is the outcome; all other columns are predictors.
recipe_lr <- recipe(grades_2324 ~ ., data = train_w) |>
  
  # Normalise all numeric predictors to mean = 0 and sd = 1.
  # Required because the Ridge penalty treats all variables equally. Without normalisation, variables on larger scales get unfairly penalised.
  step_normalize(all_numeric_predictors())

# Define the logistic regression model.
# penalty = tune() means we will search for the best regularisation strength.
# mixture = 0 means Ridge (L2) 
lr_spec <- multinom_reg(
  penalty = tune(),
  mixture = 0
) |>
  set_engine("glmnet")

# Bundle the recipe, model, and class weights into a single workflow.
# add_case_weights() tells the workflow to use the case_wt column as weights. It will NOT be used as a predictor.
lr_wf <- workflow() |>
  add_recipe(recipe_lr) |>
  add_model(lr_spec) |>
  add_case_weights(case_wt)
```

---

# Step 5: Tune the Penalty with Cross-Validation

```{r tune, message=FALSE, warning=FALSE}
# strata ensures each fold has the same class proportions as the full training set.
cv_folds <- vfold_cv(train_w, v = 5, strata = grades_2324)

# Create a random search grid of 20 penalty values
# Values are spread on a log scale (small to large regularisation strength).
set.seed(21)
penalty_grid <- grid_random(penalty(), size = 20)

# Run 5-fold CV for every penalty value.
# macro recall averages recall across all four grade classes equally.
set.seed(21)
lr_tune <- tune_grid(
  lr_wf,
  resamples = cv_folds,
  grid      = penalty_grid,
  metrics   = metric_set(recall)
)

show_best(lr_tune, metric = "recall", n = 5)
```

```{r tune-plot, fig.height=4, fig.width=6}
# Plot recall vs penalty to see how the model performs across the search space
autoplot(lr_tune) +
  labs(
    title    = "Tuning Curve: Macro Recall vs Penalty",
    subtitle = "Each point = average across 5 CV folds"
  )
```

---

# Step 6: Final Model

```{r final-fit, message=FALSE, warning=FALSE}
# Pick the penalty value that gave the highest macro recall across the CV folds
best_penalty <- select_best(lr_tune, metric = "recall")
cat("Best penalty:", round(best_penalty$penalty, 5), "\n")

# Swap tune() for the actual best penalty value
final_wf <- lr_wf |> finalize_workflow(best_penalty)

# last_fit() trains on the FULL training set and evaluates once on the test set.
# This is the only time we use test data
final_fit <- last_fit(
  final_wf,
  split   = split,
  metrics = metric_set(accuracy, recall)
)

```

---

# Step 7: Confusion Matrix

```{r conf-mat, fig.height=5, fig.width=6}
# last_fit() doesn't save predictions when case weights are used,
# so we manually fit on the full training set and predict on the test set.
fitted_final <- final_wf |> fit(data = train_w)

# predict() returns the predicted class for each row in the test set.
# We bind the true labels alongside so we can compare them.
preds <- predict(fitted_final, new_data = test) |>
  bind_cols(test |> select(grades_2324))

# Build the confusion matrix
cm <- conf_mat(preds, truth = grades_2324, estimate = .pred_class)

# Heatmap: dark = many predictions, light = few.
# Ideally the diagonal (correct predictions) should be the darkest cell in each row.
autoplot(cm, type = "heatmap") +
  labs(
    title    = "Confusion Matrix: Logistic Regression",
    subtitle = "Test set | Class-weighted training | Macro recall"
  )
```

```{r class-report}
# Per-class precision, recall, and F1
cm |>
  summary() |>
  filter(.metric %in% c("sensitivity", "precision", "f_meas")) |>
  select(.metric, .estimate) |>
  mutate(.estimate = round(.estimate, 3))
```

---

# Step 8: Feature Importance

Which variables did the model rely on most? We extract importance scores (mean absolute
coefficient across the four outcome classes).

```{r vip-plot, fig.height=6, fig.width=7, message=FALSE, warning=FALSE}
# vi() extracts a tidy table of variable importance scores.
# For glmnet, importance = mean absolute coefficient value across outcome classes.
# fitted_final is the class-weighted model trained on the full training set.
vip_data <- fitted_final |>
  extract_fit_parsnip() |>
  vi() |>
  # Keep only the top 20 most important variables to avoid a cluttered plot
  slice_max(Importance, n = 20)

# Plot importance scores as a horizontal bar chart
ggplot(vip_data, aes(x = Importance, y = reorder(Variable, Importance))) +
  geom_col() +
  labs(
    title    = "Top 20 Predictors -- Logistic Regression (Class Weights)",
    subtitle = "Importance = mean absolute coefficient across outcome classes",
    x = "Importance", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

---

# Step 9: Conclusion

## Model performance summary

```{r summary-table}
# Compute accuracy and macro recall directly from the test-set predictions.
# We use preds (manual predictions) because last_fit() does not save predictions
# when case weights are present in the workflow.
preds |>
  summarise(
    accuracy = mean(.pred_class == grades_2324),
    recall   = recall_vec(truth = grades_2324, estimate = .pred_class,
                          estimator = "macro")
  ) |>
  mutate(across(everything(), \(x) round(x, 3)))
```

## Key findings

**Food security is a meaningful predictor of grades.** Both `FoodSit_2324` and `FoodCash_2324`
appear in the top predictors after controlling for age, sex, poverty level, and family structure.
Children in households that struggle with food access are more likely to receive fair or poor grades.

**Other factors also matter.** ADHD severity (`ADHDSevInd_2324`), special education plans
(`SpEducPln_2324`), and social factors like bullying and ability to make friends are among the
stronger predictors. Grades are shaped by a combination of food access, health, and social
circumstances.

**Model limitations.** A macro recall of 46% means the model correctly identifies roughly 4-5
out of every 10 at-risk children, better than a model that ignores minority classes entirely,
but not reliable enough to use as a standalone screening tool. These are associations from
cross-sectional survey data. We cannot claim food insecurity *causes* worse grades.



# Step 10: Inferential Analysis: How Much More Likely?

The predictive model tells us *that* food security matters. This section tells us *by how much*,
using a standard (unregularised) logistic regression to produce **odds ratios** with confidence
intervals and p-values.

The predictive model uses Ridge regularisation, which deliberately shrinks coefficients to
improve prediction. Those shrunken coefficients cannot be interpreted as valid estimates of
effect size. An inferential model is designed purely to estimate relationships accurately.

## Create a binary outcome

```{r binary-outcome}
df_infer <- df_model |>
  mutate(
    # Create a binary outcome: at_risk = Fair or Poor; not_at_risk = Excellent or Good
    at_risk = factor(
      ifelse(grades_2324 %in% c("Fair", "Poor"), "at_risk", "not_at_risk"),
      levels = c("not_at_risk", "at_risk")  # "not_at_risk" is the reference category
    )
  )

# Check the binary split
df_infer |>
  count(at_risk) |>
  mutate(pct = round(n / sum(n) * 100, 1))
```

## Fit the inferential logistic regression

Food security variables are the **focal predictors**. All other variables are **controls**
included only to prevent confounding.

Note: we use `ADHDSevInd_2324` rather than `ADHDind_2324` because the two are nearly perfectly
correlated (r = 0.94). `ADHDSevInd` is preferred as it contains severity information in addition
to presence/absence -- including both would cause multicollinearity.

```{r infer-glm}
# glm() fits a standard logistic regression with no regularisation.
# family = binomial tells R the outcome is binary (at_risk vs not_at_risk).
infer_fit <- glm(
  at_risk ~ FoodSit_2324 + FoodCash_2324 +                           # food security (focal)
            age5_2324 + sex_2324 + povlev4_2324 + famstruct5_2324 +  # demographic controls
            ADHDSevInd_2324,                                          # ADHD control
  data   = df_infer,
  family = binomial
)

summary(infer_fit)
```

## Extract odds ratios

```{r odds-ratios}
# exponentiate = TRUE converts log-odds to odds ratios (easier to interpret).
# conf.int = TRUE adds 95% confidence intervals.
odds_ratios <- tidy(infer_fit, exponentiate = TRUE, conf.int = TRUE) |>

  # Remove the intercept -- it has no meaningful interpretation here
  filter(term != "(Intercept)") |>

  # Label each variable by its role
  mutate(
    role = case_when(
      term %in% c("FoodSit_2324", "FoodCash_2324")                            ~ "Food security (focal)",
      term %in% c("age5_2324", "sex_2324", "povlev4_2324", "famstruct5_2324",
                  "ADHDSevInd_2324")                                           ~ "Control variable"
    ),
    # Flag statistical significance at the 5% level
    significant = p.value < 0.05
  ) |>
  arrange(desc(estimate))

# Print results table
odds_ratios |>
  select(term, estimate, conf.low, conf.high, p.value, role) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## Visualise odds ratios

```{r or-plot, fig.height=4, fig.width=7}
# Each point is an odds ratio; horizontal lines are 95% confidence intervals.
# Dashed vertical line at OR = 1 is the "no effect" reference.
# OR > 1 = higher odds of at-risk grades, OR < 1 = lower odds.
ggplot(odds_ratios, aes(x = estimate, y = reorder(term, estimate), color = role)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c(
    "Food security (focal)" = "#2166ac",
    "Control variable"      = "#b2b2b2"
  )) +
  labs(
    title    = "Odds Ratios -- Inferential Logistic Regression",
    subtitle = "Outcome: Fair/Poor vs Excellent/Good. OR > 1 = higher odds of at-risk grades.",
    x = "Odds Ratio (95% CI)", y = NULL, color = NULL
  ) +
  theme(legend.position = "bottom")
```

## Interpretation

**Food security findings:**

- **`FoodSit_2324` (OR around 1.42):** Each step toward worse food access -- from "always enough
  nutritious food" toward "often not enough" -- is associated with approximately 42% higher odds
  of at-risk grades, after controlling for age, sex, poverty, family structure, and ADHD severity.

- **`FoodCash_2324` (OR around 1.40):** Children from families relying on more types of food and cash
  assistance(a marker of greater economic need) have approximately 40% higher odds of at-risk
  grades. Note: this variable captures *need* for assistance, not a harmful effect of receiving it.

Both results are statistically significant (p < 0.001) with tight confidence intervals.

**How to write this up:**

> *"Each step toward greater food insecurity was associated with 1.42 times higher odds of
> receiving fair or poor grades, after controlling for age, sex, poverty level, family structure,
> and ADHD severity (OR = 1.42, 95% CI [1.37, 1.48], p < 0.001)."*

**Important caveat:** These are observational associations, not causal effects. We cannot rule
out that unmeasured factors, such as neighborhood safety or parental stress, drive both food
insecurity and lower grades simultaneously.
