---
title: "Logistic Regression: Grade Outcomes"
author: "Gabrielle Arrache and Naomi Francis"
output: html_document
date: "`r Sys.Date()`"
---

# Setup
```{r setup, message=FALSE, warning=FALSE}
library(tidymodels)
library(glmnet)
library(vip)
library(broom)
```

---

# Data Preparation
run clean_data.Rmd first

---

# Data Splitting

```{r split}
set.seed(21)  


split <- initial_split(df_model, prop = 0.70, strata = grades_2324)
train <- training(split)   
test  <- testing(split) 

nrow(train)
nrow(test)
```

---

# Class Imbalance Handling

Because the lower-grade classes are less common, I used class weights so the model gives them more attention during training.

```{r class-weights}
n_total <- nrow(train)
class_counts <- table(train$grades_2324)

# Calculate weights per class
weights_lookup <- n_total / class_counts
print(round(weights_lookup, 2))

train_w <- train |>
  mutate(
    case_wt = importance_weights(
      as.numeric(weights_lookup[as.character(grades_2324)])
    )
  )
```

---

# Workflow

```{r recipe-and-model}
recipe_lr <- recipe(grades_2324 ~ ., data = train_w) |>
  step_normalize(all_numeric_predictors())


lr_spec <- multinom_reg(
  penalty = tune(),
  mixture = 0
) |>
  set_engine("glmnet")

lr_wf <- workflow() |>
  add_recipe(recipe_lr) |>
  add_model(lr_spec) |>
  add_case_weights(case_wt)
```

---

# Tuning

We optimize the penalty parameter using 5-fold cross-validation, with macro recall as our primary metric.

```{r tune, message=FALSE, warning=FALSE}
cv_folds <- vfold_cv(train_w, v = 5, strata = grades_2324)

set.seed(21)
penalty_grid <- grid_random(penalty(), size = 20)

set.seed(21)
lr_tune <- tune_grid(
  lr_wf,
  resamples = cv_folds,
  grid      = penalty_grid,
  metrics   = metric_set(recall)
)

show_best(lr_tune, metric = "recall", n = 5)
```

```{r tune-plot, fig.height=4, fig.width=6}
autoplot(lr_tune) +
  labs(
    title    = "Tuning Curve: Macro Recall vs Penalty",
    subtitle = "Averaged across 5-fold CV"
  )
```

---

# Fit

```{r final-fit, message=FALSE, warning=FALSE}
best_penalty <- select_best(lr_tune, metric = "recall")
cat("Optimal penalty:", round(best_penalty$penalty, 5), "\n")

final_wf <- lr_wf |> finalize_workflow(best_penalty)

final_fit <- last_fit(
  final_wf,
  split   = split,
  metrics = metric_set(accuracy, recall)
)
```

---

# Model Evaluation

```{r conf-mat, fig.height=5, fig.width=6}
# Manual fit to extract predictions (handling case weight limitations in last_fit)
fitted_final <- final_wf |> fit(data = train_w)

preds <- predict(fitted_final, new_data = test) |>
  bind_cols(test |> select(grades_2324))

cm <- conf_mat(preds, truth = grades_2324, estimate = .pred_class)

autoplot(cm, type = "heatmap") +
  labs(
    title    = "Confusion Matrix",
    subtitle = "Test set performance (Class-weighted)"
  )
```

```{r class-report}
cm |>
  summary() |>
  filter(.metric %in% c("sensitivity", "precision", "f_meas")) |>
  select(.metric, .estimate) |>
  mutate(.estimate = round(.estimate, 3))
```

---

# Variable Importance

```{r vip-plot, fig.height=6, fig.width=7, message=FALSE, warning=FALSE}
vip_data <- fitted_final |>
  extract_fit_parsnip() |>
  vi() |>
  slice_max(Importance, n = 20)

ggplot(vip_data, aes(x = Importance, y = reorder(Variable, Importance))) +
  geom_col() +
  labs(
    title    = "Top 20 Predictors",
    subtitle = "Based on Ridge regression importance scores",
    x = "Importance", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

---

# Summary of Findings

## Performance

```{r summary-table}
preds |>
  summarise(
    accuracy = mean(.pred_class == grades_2324),
    recall   = recall_vec(truth = grades_2324, estimate = .pred_class,
                          estimator = "macro")
  ) |>
  mutate(across(everything(), \(x) round(x, 3)))
```


---

# Inferential Analysis: Odds Ratios

To better understand the magnitude of these effects, we use an unregularized logistic regression on a binary outcome (At Risk vs. Not At Risk) to calculate Odds Ratios.

## Binary Outcome

```{r binary-outcome}
df_infer <- df_model |>
  mutate(
    at_risk = factor(
      ifelse(grades_2324 %in% c("Fair", "Poor"), "at_risk", "not_at_risk"),
      levels = c("not_at_risk", "at_risk")
    )
  )

df_infer |>
  count(at_risk) |>
  mutate(pct = round(n / sum(n) * 100, 1))
```

## Model Estimation

We focused on food security variables while controlling for age, sex, poverty, family structure, and ADHD severity.

```{r infer-glm}
infer_fit <- glm(
  at_risk ~ FoodSit_2324 + FoodCash_2324 +                           
            age5_2324 + sex_2324 + povlev4_2324 + famstruct5_2324 +  
            ADHDSevInd_2324,                                          
  data   = df_infer,
  family = binomial
)

summary(infer_fit)
```

## Results

```{r odds-ratios}
odds_ratios <- tidy(infer_fit, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    role = case_when(
      term %in% c("FoodSit_2324", "FoodCash_2324") ~ "Food security (focal)",
      TRUE ~ "Control variable"
    ),
    significant = p.value < 0.05
  ) |>
  arrange(desc(estimate))

odds_ratios |>
  select(term, estimate, conf.low, conf.high, p.value, role) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## Visualizing Effects

```{r or-plot, fig.height=4, fig.width=7}
ggplot(odds_ratios, aes(x = estimate, y = reorder(term, estimate), color = role)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c(
    "Food security (focal)" = "#2166ac",
    "Control variable"      = "#b2b2b2"
  )) +
  labs(
    title    = "Effect Estimates (Odds Ratios)",
    subtitle = "Outcome: Fair/Poor vs Excellent/Good grades",
    x = "Odds Ratio (95% CI)", y = NULL, color = NULL
  ) +
  theme(legend.position = "bottom")
```


## Saving Model
```{r} 
saveRDS(fitted_final, "Logist_Regression_model.rds")
```
