---
title: "single_factor_models"
output: html_document
date: "2026-04-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyzing single factor models to determine which models can explain grades most

```{r}
ChHlthSt_model <- lm(grades_2324 ~ 1 +ChHlthSt_2324, data = df_23_24_clean)
ChHlthSt_R <- summary(ChHlthSt_model)$r.squared 

LowBWght_model <- lm(grades_2324 ~ 1 +LowBWght_2324, data = df_23_24_clean)
LowBWght_R <- summary(LowBWght_model)$r.squared 

PhysAct_model <- lm(grades_2324 ~ 1 +PhysAct_2324, data = df_23_24_clean)
PhysAct_R <- summary(PhysAct_model)$r.squared 

Cond2more_model <- lm(grades_2324 ~ 1 + Cond2more_2324, data = df_23_24_clean)
Cond2more_R <- summary(Cond2more_model)$r.squared

CSHCN_model <- lm(grades_2324 ~ 1 + CSHCN_2324, data = df_23_24_clean)
CSHCN_R <- summary(Cond2more_model)$r.squared

ADHDSevInd_model <- lm(grades_2324 ~ 1 + ADHDSevInd_2324, data = df_23_24_clean)
ADHDSevInd_R <- summary(ADHDSevInd_model)$r.squared

bullied_model <- lm(grades_2324 ~ 1 + bullied_2324, data = df_23_24_clean)
bullied_R <- summary(bullied_model)$r.squared

bully_model <- lm(grades_2324 ~ 1 + bully_2324, data = df_23_24_clean)
bully_R <- summary(bully_model)$r.squared

MakeFriend_model <- lm(grades_2324 ~ 1 + MakeFriend_2324, data = df_23_24_clean)
MakeFriend_R <- summary(MakeFriend_model)$r.squared

MEDB10ScrQ5_model <- lm(grades_2324 ~ 1 + MEDB10ScrQ5_2324, data = df_23_24_clean)
MEDB10ScrQ5_R <- summary(MEDB10ScrQ5_model)$r.squared

povlev4_model <- lm(grades_2324 ~ 1 + povlev4_2324, data = df_23_24_clean)
povlev4_R <- summary(povlev4_model)$r.squared

FoodSit_model <- lm(grades_2324 ~ 1 + FoodSit_2324, data = df_23_24_clean)
FoodSit_R <- summary(FoodSit_model)$r.squared

HousingInstab_model <- lm(grades_2324 ~ 1 + HousingInstab_2324, data = df_23_24_clean)
HousingInstab_R <- summary(HousingInstab_model)$r.squared

famstruct5_model <- lm(grades_2324 ~ 1 + famstruct5_2324, data = df_23_24_clean)
famstruct5_R <- summary(famstruct5_model)$r.squared

age5_model <- lm(grades_2324 ~ 1 + age5_2324, data = df_23_24_clean)
age5_R <- summary(age5_model)$r.squared

sex_model <- lm(grades_2324 ~ 1 + sex_2324, data = df_23_24_clean)
sex_R <- summary(sex_model)$r.squared

MedCare_model <- lm(grades_2324 ~ 1 + MedCare_2324, data = df_23_24_clean)
MedCare_R <- summary(MedCare_model)$r.squared

MentHCare_model <- lm(grades_2324 ~ 1 + MentHCare_2324, data = df_23_24_clean)
MentHCare_R <- summary(MentHCare_model)$r.squared

SpEducPln_model <- lm(grades_2324 ~ 1 + SpEducPln_2324, data = df_23_24_clean)
SpEducPln_R <- summary(SpEducPln_model)$r.squared

ReptGrade_model <- lm(grades_2324 ~ 1 + ReptGrade_2324, data = df_23_24_clean)
ReptGrade_R <- summary(ReptGrade_model)$r.squared

```

## Combine into 1 dataframe
```{r}
library(tidyr)

all_r_squared <- data.frame(ChHlthSt_R, LowBWght_R, PhysAct_R, Cond2more_R, CSHCN_R, ADHDSevInd_R, bullied_R, bully_R, MakeFriend_R, MEDB10ScrQ5_R, povlev4_R, FoodSit_R, HousingInstab_R, famstruct5_R, age5_R, sex_R, MedCare_R, MentHCare_R, SpEducPln_R, ReptGrade_R)

all_r_squared <- all_r_squared |>
  pivot_longer(cols = 1:20, names_to ="factor", values_to = "R squared")

```

After ordering factors by R squared value, the most influential ones are 
1) Mental, emotional, developmental, or behavioral problems(MEDB10ScrQ5) 
2) ADHD Severity Index (ADHDSevInd) 
3) Special Education Plan Status (SpEducPln) 
4) Health Conditions (Cond2more) 
5) Special Health Care Needs (CSHCN)
6) Difficulty making/keeping friends (MakeFriend)
7) Food Situation (FoodSit)
8) Poverty Level (povlev4)

Making a model with these 8 most influential variables:
```{r}
top_eight_model <- lm(grades_2324 ~ 1 + MEDB10ScrQ5_2324 + ADHDSevInd_2324 + SpEducPln_2324 + Cond2more_2324 + CSHCN_2324 + MakeFriend_2324 + FoodSit_2324 + povlev4_2324, data = df_23_24_clean)

summary(top_eight_model)$r.squared
```


