---
title: "Three_profiles"
output: html_document
date: "2026-05-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Profile 1
### 10 year old female with rated moderate autism. This may represent a student who is on the autism spectrum and has some behavior problems, but has the resources and family support to help with school. On a special education plan, married parents, not living in poverty, physically active, not food insecure, very good health, low birth weight, currently has condition(cond2more), four or five screener criteria, does not have adhd, bullied 1 or 2 times per month, does not bully others, a lot of difficulty making friends, has gotten a physical checkup in the past year, has not recieved mental health care, has not repeated a grade, always gets enough nutritious food, did not experience housing instability, 

```{r}
#building the profile
profile_1 <- data.frame(
      SpEducPln_2324 = 1, # on a special ed plan
      famstruct5_2324 = 1,  # married parents
      povlev4_2324 = 4, # over 400% poverty
      PhysAct_2324 = 1, # gets 60 mins a day
      FoodCash_2324 = 1, # does not recieve food cash
      sex_2324 = 2, # female
      age5_2324 = 3, # 8-11 yrs old
      ChHlthSt_2324 = 1, # excellent or very good
      LowBWght_2324 = 1, # yes
      Cond2more_2324 = 3, # currently has condition
      CSHCN_2324 = 3, # three screener criteria
      ADHDind_2324 = 1, # does not have condition
      ADHDSevInd_2324 = 1, # does not have condition
      bullied_2324 = 3, # bullied once or twice a month
      bully_2324 = 1, # never bully others in past 12 months
      MakeFriend_2324 = 3, # a lot of difficulty
      MEDB10ScrQ5_2324 = 3, # currently has condition
      MedCare_2324 = 1, #yes
      MentHCare_2324 = 3, # no need
      ReptGrade_2324 = 2, # no
      FoodSit_2324 = 1, # always get enough nutritious food
      HousingInstab_2324 = 2, # did not experience housing insecurity
      ASDSevInd_2324 = 3, # ASD rated moderate or severe
      case_wt = 1)
```

## predicting the profile 1 grades using the logistic model `fitted_final`
```{r}
profile_1_pred <-  predict(fitted_final, new_data = profile_1)
profile_1_pred
```
### Profile 1 is predicted to have good grades.

## Profile 2
### A 16 year old male with mild ADHD who has a single parent with low income, and lives with 2 other siblings. This represents a student who is food insecure. The student bullies others occasionally (once or twice a month). He did experence housing instability
```{r}
#building the profile
profile_2 <- data.frame(
      SpEducPln_2324 = 2, # not on a special ed plan
      famstruct5_2324 = 3,  # single parent
      povlev4_2324 = 2, # 100 - 199 percent poverty level
      PhysAct_2324 = 2, # doesn't get 60 mins a day
      FoodCash_2324 = 2, # one or two types
      sex_2324 = 1, # male
      age5_2324 = 5, # 15-17 yrs old
      ChHlthSt_2324 = 2, # good
      LowBWght_2324 = 2, # no
      Cond2more_2324 = 1, # does not have condition
      CSHCN_2324 = 1, # one condition
      ADHDind_2324 = 3, # currently has condition
      ADHDSevInd_2324 = 2, # current condition, rated mild
      bullied_2324 = 1, # never been bullied in past 12 months
      bully_2324 = 3, # bullies other once or twice a month
      MakeFriend_2324 = 1, # no difficulty
      MEDB10ScrQ5_2324 = 3, # currently has condition
      MedCare_2324 = 2, # none in past year
      MentHCare_2324 = 3, # no need
      ReptGrade_2324 = 2, # no
      FoodSit_2324 = 3, # sometimes not enough food
      HousingInstab_2324 = 1, # did  experience housing insecurity
      ASDSevInd_2324 = 1, # does not have ASD
      case_wt = 1)
```

## predicting the profile 2 grades using the logistic model `fitted_final`
```{r}
profile_2_pred <-  predict(fitted_final, new_data = profile_2)
profile_2_pred
```
### Profile 2 is predicted to have good grades.

## Profile 3 is a 6 year old student who has no reported issues with mental or physical health. However, he has trouble making friends at school and is bullied every day. He occasionally bullies others, too.
```{r}
#building the profile
profile_3 <- data.frame(
      SpEducPln_2324 = 2, # not on a special ed plan
      famstruct5_2324 = 2,  # not married parents
      povlev4_2324 = 4, # 400%+ over poverty level
      PhysAct_2324 = 1, # does get 60 mins a day exercise
      FoodCash_2324 = 1, # none
      sex_2324 = 1, # male
      age5_2324 = 2, # 4-7 yrs old
      ChHlthSt_2324 = 1, # very good
      LowBWght_2324 = 2, # no
      Cond2more_2324 = 1, #  no condition
      CSHCN_2324 = 1, # one criteria of special needs
      ADHDind_2324 = 1, # no adhd
      ADHDSevInd_2324 = 1, # no condition
      bullied_2324 = 5, # every day
      bully_2324 = 3, # twice a month
      MakeFriend_2324 = 3, # a lot of difficulty
      MEDB10ScrQ5_2324 = 1, # does not have condition
      MedCare_2324 = 1, # yes, check up in past year
      MentHCare_2324 = 3, # no need
      ReptGrade_2324 = 2, # no
      FoodSit_2324 = 1, # always enough food
      HousingInstab_2324 = 1, # did not experience housing insecurity
      ASDSevInd_2324 = 1, # no ASD
      case_wt = 1)
```

```{r}
profile_3_pred <-  predict(fitted_final, new_data = profile_3)
profile_3_pred
```
