---
title: "Logistic Regression: Grade Outcomes"
author: "Gabrielle Arrache and Naomi Francis"
output: html_document
date: "`r Sys.Date()`"
---

# Setup
```{r setup, message=FALSE, warning=FALSE}
library(tidymodels)
library(glmnet)
library(vip)
library(broom)
```

---

# Data Preparation
```{r}
#run clean_data.Rmd (data folder) first
df_model <- df_23_24_clean |>
  mutate(grades_2324 = factor(
    grades_2324, 
    levels = c(1, 2, 3, 4),
    labels = c("Excellent", "Good", "Fair", "Poor")
  ))


cat("Grade distribution (%):\n")
round(prop.table(table(df_model$grades_2324)) * 100, 1)
```


---

# Data Splitting

```{r split}
set.seed(21) #making code reproducible

# split the data into training and test datasets
# 70% is used to train the model
# 30% is used to test how well the model performs

split <- initial_split(df_model, prop = 0.70, strata = grades_2324)
train <- training(split)   
test  <- testing(split) 

nrow(train)
nrow(test)
```

---

# Class Imbalance Handling

Because the lower-grade classes are less common, so we used class weights so the model gives them more attention during training.

```{r class-weights}
n_total <- nrow(train)
class_counts <- table(train$grades_2324)

# Calculate weights per class
weights_lookup <- n_total / class_counts
print(round(weights_lookup, 2))

train_w <- train |>
  mutate(
    case_wt = importance_weights(
      as.numeric(weights_lookup[as.character(grades_2324)])
      )
  )
```

---

# Workflow

```{r recipe-and-model}
recipe_lr <- recipe(grades_2324 ~ ., data = train_w) |>
  step_normalize(all_numeric_predictors()) #Normalize numeric variables, which puts all numbers on a similar scale

lr_spec <- multinom_reg( #build the logistic regression model
  penalty = tune(), #the penalty helps reduce overfitting
  mixture = 0) |>
  set_engine("glmnet")

lr_wf <- workflow() |> # combine the recipe and the model into one workflow
  add_recipe(recipe_lr) |>
  add_model(lr_spec) |>
  add_case_weights(case_wt)
```

---

# Tuning

We optimize the penalty parameter using 5-fold cross-validation, with macro recall as our primary metric.

```{r tune, message=FALSE, warning=FALSE}
cv_folds <- vfold_cv(train_w, v = 5, strata = grades_2324) #create 5 different folds for cross-validation to find the best version

set.seed(21)
penalty_grid <- grid_random(penalty(), size = 20) #try different penalty values to find which one gives the best recall

set.seed(21)
lr_tune <- tune_grid(
  lr_wf,
  resamples = cv_folds,
  grid      = penalty_grid,
  metrics   = metric_set(recall)
)

show_best(lr_tune, metric = "recall", n = 5) 
```

```{r tune-plot, fig.height=4, fig.width=6}
autoplot(lr_tune) +
  labs(
    title    = "Tuning Curve: Macro Recall vs Penalty: Averaged across 5-fold CV")
```

---

# Fit

```{r final-fit, message=FALSE, warning=FALSE}
best_penalty <- select_best(lr_tune, metric = "recall") #select the best penalty value
cat("Optimal penalty:", round(best_penalty$penalty, 5), "\n")

final_wf <- lr_wf |> finalize_workflow(best_penalty)

final_fit <- last_fit( #fit the final model using the best settings
  final_wf,
  split   = split,
  metrics = metric_set(accuracy, recall))
```

---

# Model Evaluation

```{r conf-mat, fig.height=5, fig.width=6}
# Manual fit to extract predictions (handling case weight limitations in last_fit)
fitted_final <- final_wf |> fit(data = train_w)

preds <- predict(fitted_final, new_data = test) |> #generate predictions on the test data to compare predicted grades with real grades
  bind_cols(test |> select(grades_2324))

cm <- conf_mat(preds, truth = grades_2324, estimate = .pred_class)

autoplot(cm, type = "heatmap") + #create a confusion matrix, which shows where the model predicted correctly and the mistakes made
  labs(
    title    = "Confusion Matrix")
```

```{r class-report}
# Measure model performance:  Precision = how often predictions are correct, recall = how many real cases the model found, F1-score = balance between precision and recall

cm |>
  summary() |>
  filter(.metric %in% c("sensitivity", "precision", "f_meas")) |>
  select(.metric, .estimate) |>
  mutate(.estimate = round(.estimate, 3))
```

---

# Variable Importance

```{r vip-plot, fig.height=6, fig.width=7, message=FALSE, warning=FALSE}

# Show the most important variables in the model
vip_data <- fitted_final |>
  extract_fit_parsnip() |>
  vi() |>
  slice_max(Importance, n = 20)

ggplot(vip_data, aes(x = Importance, y = reorder(Variable, Importance))) +
  geom_col() +
  labs(
    title    = "Top 20 Predictors based on Ridge regression importance scores",
    x = "Importance", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom") +
  theme_classic()
```

---

# Summary of Findings

## Performance

```{r summary-table}
# create a summary table for the accuracy and recall
preds |>
  summarise(
    accuracy = mean(.pred_class == grades_2324), # correct predictions out of all predictions
    recall   = recall_vec(truth = grades_2324, estimate = .pred_class, # how many real cases the model found
                          estimator = "macro")
  ) |>
  mutate(across(everything(), \(x) round(x, 3)))
```


---

# Inferential Analysis: Odds Ratios

To better understand the magnitude of these effects, we use an unregularized logistic regression on a binary outcome (At Risk vs. Not At Risk) to calculate Odds Ratios.

## Binary Outcome

```{r binary-outcome}
# create two outcomes, "at risk" and "not at risk"
df_infer <- df_model |>
  mutate(
    at_risk = factor(
      ifelse(grades_2324 %in% c("Fair", "Poor"), "at_risk", "not_at_risk"),
      levels = c("not_at_risk", "at_risk")
    )
  )

# count how many are in each category
df_infer |>
  count(at_risk) |>
  mutate(pct = round(n / sum(n) * 100, 1))
```

## Model Estimation

We focused on food security variables while controlling for age, sex, poverty, family structure, and ADHD severity.

```{r infer-glm}
# use the top influential variables identified to predict at risk or not at risk grades
infer_fit <- glm(
  at_risk ~ FoodSit_2324 + FoodCash_2324 +                           
            age5_2324 + sex_2324 + povlev4_2324 + famstruct5_2324 +  
            ADHDSevInd_2324,                                          
  data   = df_infer,
  family = binomial # two outcomes
)

# summary of the model
summary(infer_fit)
```

## Results

```{r odds-ratios}
# compute odds ratios of children having at risk grades for each factor in the model
odds_ratios <- tidy(infer_fit, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    role = case_when(
      term %in% c("FoodSit_2324", "FoodCash_2324") ~ "Food security (focal)",
      TRUE ~ "Control variable"
    ),
    significant = p.value < 0.05
  ) |>
  arrange(desc(estimate))

# show the odds of a student haveing at risk grades in each category
odds_ratios |>
  select(term, estimate, conf.low, conf.high, p.value, role) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## Visualizing Effects

```{r or-plot, fig.height=4, fig.width=7}
# treating food security variables, FoodSit and FoodCash, as the focal points, look at the odds in comparison to other influential variables

ggplot(odds_ratios, aes(x = estimate, y = reorder(term, estimate), color = role)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c(
    "Food security (focal)" = "#2166ac",
    "Control variable"      = "#b2b2b2"
  )) +
  labs(
    title    = "Effect Estimates (Odds Ratios)",
    subtitle = "Outcome: Fair/Poor vs Excellent/Good grades",
    x = "Odds Ratio (95% CI)", y = NULL, color = NULL
  ) +
  theme(legend.position = "bottom")
```


## Blog Summary Statistics
```{r blog-stats}

# create statistics about the data for the blog
blog_stats <- list(
  n_children        = scales::comma(nrow(df_model)),
  pct_food_insecure = scales::percent(mean(df_model$FoodSit_2324 %in% c(3, 4)), accuracy = 0.1),
  pct_at_risk       = scales::percent(mean(df_infer$at_risk == "at_risk"),        accuracy = 0.1),
  pct_food_assist   = scales::percent(mean(df_model$FoodCash_2324 %in% c(2, 3)), accuracy = 0.1)
)
```

## Prepare Blog Prediction Objects
```{r blog-preds}
# preparing for using the model in the blog
test_preds <- preds

test_probs <- predict(fitted_final, new_data = test, type = "prob") |>
  bind_cols(test |> select(grades_2324))
```

## Saving Model
```{r save-all} 
# save models in models folder 
saveRDS(fitted_final,  "../models/Logist_Regression_model.rds")
saveRDS(fitted_final,  "../models/fitted_final.rds")
saveRDS(odds_ratios,   "../models/odds_ratios.rds")
saveRDS(vip_data,      "../models/vip_data.rds")
saveRDS(test_preds,    "../models/test_preds.rds")
saveRDS(test_probs,    "../models/test_probs.rds")
saveRDS(blog_stats,    "../models/blog_stats.rds")
```
